---
output: 
  pdf_document:
    fig_width: 6
    fig_height: 4
fontsize: 10pt
papersize: a4paper
header-includes:
- \usepackage{amsmath,amsfonts}
---

\textbf{STATISTICAL MODELLING} \hfill Part IIC, Michaelmas 2019

\vspace{-0.2cm}
\textbf{Practical 4: More on linear regression} \hfill By courtesy of Dr. S. A. Bacallado and Dr. R. Shah

## Solutions

```{r}
file_path <- "https://raw.githubusercontent.com/AJCoca/SM19/master/"
Movies <- read.csv(paste(file_path, "Movies.csv", sep =""))
attach(Movies)

MoviesLM <- lm(log(Total.Gross) ~ log(Opening) + Screens + RT + log(Budget))
lev <- hatvalues(MoviesLM)
high_lev <- which(lev > 3*4/nrow(Movies)) # gives the high leverage obs
lev[high_lev] # actually shows their leverage
```

Observation 99 has much higher leverage than the rest. This is largely because the budget was so low for the film.
```{r}
Movies[99, ]
```

We now fit a model without this observation
```{r}
MoviesLM_sub <- lm(log(Total.Gross) ~ log(Opening) +  RT + log(Budget), subset=-99)


## Prediction intervals
Movies2010 <- read.csv(paste(file_path, "Movies2010.csv", sep =""))

pred_intervals <- predict(MoviesLM_sub, Movies2010, interval="prediction")
pred_intervals_trans <- exp(pred_intervals)

target <- Movies2010$Total.Gross
mean((pred_intervals_trans[, 2] < target) & (pred_intervals_trans[, 3] > target))
# (pred_intervals_trans[, 2] < target) gives a logical vector
# The & performs a copmonentwise AND operation
# Applying the mean function first coerces the logical vector into an vector where
# TRUE -> 1 and FALSE -> 0

```

Every film is inside its prediction interval. We can also see how well our model predicts film earnings by displaying the true earnings alongside the predicted earnings.

```{r}
cbind(target, pred_intervals_trans)
```

Finally, we repeat the procedure with 50% prediction intervals.

```{r}
pred_intervals <- predict(MoviesLM_sub, Movies2010, interval="prediction",level=0.5)
pred_intervals_trans <- exp(pred_intervals)

target <- Movies2010$Total.Gross
mean((pred_intervals_trans[, 2] < target) & (pred_intervals_trans[, 3] > target))
# (pred_intervals_trans[, 2] < target) gives a logical vector
# The & performs a copmonentwise AND operation
# Applying the mean function first coerces the logical vector into an vector where
# TRUE -> 1 and FALSE -> 0
```

In this case only half of the observations are in their prediction interval.